---
title: "Assignment 3: Data Exploration"
author: "Jessica Sheldon, Section #2"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name, Section #" on line 3 (above) with your name and section number.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "FirstLast_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. **Be sure to add the `stringsAsFactors = TRUE` parameter to the function when reading in the CSV files.**
```{r,  message = FALSE}
getwd() #check working directory
library(tidyverse) #load tidyverse package
Neonics <- read.csv("../Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE) #import ecotox dataset
Litter <- read.csv("../Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE) #import Litter dataset
```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Neonicotinoids are an insecticide useful against sap-feeding insects. 

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: 

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
1. Sampling for litter and woody debris only occurs at NEON sites with woody vegitation more than 2 meters tall. 
2. Sampling is completed through tower plots. 
3. Ground traps are sampled once per year.


## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}
dim(Neonics) #dimensions of the Neonics dataset
```

6. Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}
summary(Neonics$Effect) #summary of the Effect column in the Neonics data
```

> Answer: This pulls out the effects 

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}
summary(Neonics$Species.Common.Name) #summary of the common name column in Neonics data
```

> Answer: The six most commonly studied species are Honey Bee, Parasitic Wasp, Buff Tailed Bumblebee, Carniolan Honey Bee, Bumble Bee, Italian Honeybee. These are all types of stinging Hymenoptera, especially types of bees. These might be of interest over other insects because the general bee population is on the decline.

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
summary(Neonics$Conc.1..Author.)
```

> Answer:



## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r}
ggplot(Neonics) + 
  geom_freqpoly(aes(x = Publication.Year), bins = 5) +
  geom_freqpoly(aes(x = Publication.Year), bins = 5, color = "darkgray") +
  geom_freqpoly(aes(x = Publication.Year), bins = 5,  lty = 2) +
  scale_x_continuous(limits = c(1980, 2020))
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}
ggplot(Neonics) + 
  geom_freqpoly(aes(x = Test.Location)) +
  geom_freqpoly(aes(x = Test.Location), color = "darkgray") +
  geom_freqpoly(aes(x = Test.Location)) lty = 2)
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r}
ggplot(Neonics, aes(x = Endpoint)) +
  geom_bar() #create bar graph of endpoint counts

```

> Answer: 
>The 2 most common endpoints are LOEL and NOEL. 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}
class(Litter$collectDate) #determine class of collectDate, is factor not date. must change.
Litter$collectDate <-as.Date(Litter$collectDate, "%m/%d/%y") #assign dates
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique(Litter$plotID) #number of plots sampled via unique
summary(Litter$plotID) #number of plots via summary
```

> Answer: The difference is unique exclusivley provides the number of plots, while summary provides the number of times each plot is used.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r, fig.height = 3, fig.width = 4}
  ggplot(Litter, aes(x = functionalGroup)) +
  geom_bar() #create functional group box plot
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}
#
ggplot(Litter) +
  geom_boxplot(aes(x = dryMass, y = functionalGroup, group = cut_width(functionalGroup, 1))) #create functional group box plot
#
ggplot(Litter) +
  geom_violin(aes(x = dryMass, y = functionalGroup), 
              draw_quantiles = c(0.25, 0.5, 0.75)) 
```

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The boxplot provides more information about the spread than the violin plot

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: Needles and mixed litter. 